Simulating High-Dimensional Multivariate Data

### 2020-10-25 010-background.Rmd
Revise motivating example

## Motivating example: RNA-seq data from breast cancer tumor samples

Simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
Often researchers posit a negative-binomial (NB) model as counts are often over-dispersed that a Poisson model would suggest.
Yet due to inherent biological processes, gene expression data exhibits correlation (co-expression) across genes [@BE07; @Schissler2018].

@@

```{r ch010-realDataTabl, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
brca <- readRDS(file = './data/brca99.rds')
set.seed(10192020)
exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
saveRDS(exampleGenes, file = './data/exampleGenes.rds')
kable(round(brca[1:4, exampleGenes]))
## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

Figure X.

```{r ch010-realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Real data for the 3 selected high-expressing genes'}
## Describe and plot real data
GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

@@

Sklar's Theorem (ref) guarantee's that any random vector with computational feasible inverse CDFs (P4 above) and obtainable correlation matrix (within the Frechet bounds) can be obtained via transformations involving copula functions. Namely, to simulate a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, we can construct a Gaussian copula ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$. When an $Y_i$ is discrete, some care must be taken to define $F_i$. Letting

@@

Simulation of a general multivariate random vector ${\bf T}$ with margins $F_i$ based on this copula is quite simple. Ensuring that one obtains a certain dependence among the marginal distributions, however, proves challenging in our proposed Gaussian-copula-based scheme. Two core issues: 1) marginal characteristics induce bounds on the possible bivariate correlations and 2) monotone transformations deform the Pearson correlation which no closed form expression exists in general (ref? Chen2001). The Frechet bounds are well-known.

>>>>

## Motivating example: RNA-seq data

Simulating high-dimensional, non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
In humans, this results in count data corresponding to over 20,000 genes (coding genomic regions) or even over 77,000 isoforms when alternating spliced mRNA are counted.
Yet due to inherent biological processes, gene expression data exhibits correlation (co-expression) across genes [@BE07; @Schissler2018].

Specifically, we'll illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas; TCGA; see Acknowledgements).
For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
Further, to illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
Table \@ref(tab:ch010-realDataTab) displays counts for three selected genes for the first 5 patients' breast tumor samples.

@@

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
brca <- readRDS(file = './data/brca99.rds')
set.seed(10192020)
exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
saveRDS(exampleGenes, file = './data/exampleGenes.rds')
kable(round(head( brca[, exampleGenes], n = 5)),
      booktabs = TRUE,
      caption = 'Counts of 3 high-expressing genes for the first 5 observations of the BRCA dataset.'
)
## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

Figure \@ref(fig:ch010-realDataFig) plots the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).

```{r ch010-realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's $rho$ for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
## Describe and plot real data
GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

@@

Sklar's Theorem (ref) guarantees that given inverse CDFs (quantile) and obtainable correlation matrix (within the Frechet bounds) can be obtained via transformations involving copula functions. 
Namely, to simulate a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, we can construct a Gaussian copula ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$. 
When an $Y_i$ is discrete, some care must be taken to define $F_i$. Letting

@@

Simulation of a multivariate random vector ${\bf T}$ with margins $F_i$ based on this copula

### 2020-10-25 050-examples.Rmd
Revise motivating example

The RNA-seq data generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
Since this is a counting processing with no upper bound, many modeling approaches discrete random variables with infinite support.
Often the counts exhibit over-dispersion and so the negative binomial arises as a sensible model for the expression levels (\emph{gene counts}).
Moreover, the counts are correlated (co-expressed) and cannot be assumed to behave independently.
RNA-seq platforms quantify the entire transcriptome in one experimental run, resulting in high dimensional data.
In humans, this results in count data corresponding to over 20,000 genes (coding genomic regions) or even over 77,000 isoforms when alternating spliced mRNA are counted.
	This suggests simulating high-dimensional multivariate NB with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.

@@

In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters.
Our goal is to replicate the structure of a breast cancer data set (BRCA data set from The Cancer Genome Atlas).
For simplicity of illustration, we begin by filtering to retain the top 5\% highest expressing genes of the 20,501 gene measurements from $N=1212$ patients' tumor samples, resulting in $d=1026$ genes.

>>>>

@@

In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters.
Our goal is to replicate the structure of a breast cancer data set (BRCA data set from The Cancer Genome Atlas).
All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
Often researchers posit a negative binomial (NB) model as counts are often over-dispersed that a Poisson model would suggest.
This suggests simulating high-dimensional multivariate NB (MVB) with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
Notably, the $\hat{p}_i's$ are small --- ranging in $[3.934 \times 10^{-6} , 1.217 \times 10^{-2}]$.
To complete the simulation algorithm inputs, we estimate the Pearson correlation matrix $\bf{R_Y}$ and set that as the target correlation.
